\Large Exponential Random Graph Models with Big Networks: Maximum Pseudolikelihood Estimation and the\\Parametric Bootstrap

### 2017-06-26 TeX/paper.tex
Add files via upload

Even though the MCMLE is in general favored over the MPLE method there are also cases where the MPLE comes in handy. Foremost, MPLE is quick and simple, since estimation can be done by basic logistic regression and does not exert elaborate MCMC methods. As mentioned in the previous chapter the MPLE approaches the MLE as the size of the networks increase and as a consequence, is a consistent estimator (see Lindsay \cite{Lindsay1988}, Strauss and Ikeda \cite{StraussIkeda1990}, Hyvarinen \cite{Hyvarinen2006}, Desmarais and Cranmer \cite{Desmarais.2012}). This implies that for an increasing number of nodes the MPLE converges in probability to the MLE, meaning that for large enough networks the MPLE becomes an option to computationally intensive MCMC methods.\\

@@

The way we use the term \textit{consistency} in this paper is different and aligns with the way consistency is seen by Lindsay \cite{Lindsay1988}, i.e. instead of considering sub-networks, we argue that both, the MLE as well as the MPLE, approach the true coefficient values as the size of networks increases. \\%This means that the MPLE of a network on $1000$ nodes is closer to the true coefficient as the MPLE of a network on $100$ nodes. \\
\textcolor{red}{Another major advantage of the MCMLE method is, as shown by van Duijn et al. \cite{vanDuijnetal2009}, that the MCMLE is a more efficient estimator than the MPLE, meaning that the variance of the MCMLE is in general lower than the variance of the MPLE.-Rewrite} Nevertheless, as shown by Desmarais and Cranmer \cite{Desmarais.2012} the MPLE outperforms the MCMLE if the number of simulated networks used to approximate the likelihood is not chosen large enough. It is even more remarkable that the number of simulated networks needed for the MCMLE, in order to surpass the MPLE increases as the size of the network increases. This means that for very large networks it becomes difficult to determine the number of simulated networks needed in order for the MCMLE to outperform the MPLE. In other words, the larger the network of interest is, the larger one has to choose the number of simulated networks in order to justify the time intensive MCMLE approach.\\[0.3cm]
\noindent In order to demonstrate this disadvantage of the MCMLE we conduct a simulation study using Goodreau's Faux Mesa High School data, which represents a simulation of an in-school friendship network among 203 students as well as the Faux Magnolia High School data, representing an in-school friendship network among 1451 students. The data for both networks originates from Resnick et al \cite{Resnicketal1997}.\\[0.3cm]
For both networks we first calculate the MCMLE and treat the estimated coefficients as the network's true values $\theta$. For both networks we take the same parametrization, using the number of edges, the nodal attribute for gender and the geometrically weighted edgewise shared partners (gwesp)(see Hunter \cite{Hunter.2006}) distribution were we fix the decay parameter $\lambda$ at $0.25$.\\
We simulate 500 new networks using the 'true' coefficients and estimate the MPLE as well as the MCMLE of these simulated networks. For every single simulated network the MCMLE calculation is being repeated several times using 1500, 800, 400, 200, 100, 50 and 25 simulated networks in order to approximate the likelihood.

@@

$$RMSE = \sqrt{\sum_{i=1}^{500}(\theta - \hat{\theta}_i)^2}$$

@@

\begin{figure}[h]
\begin{center}
\includegraphics[scale=.5]{RMSE_paper}
\caption{The log of the ratio of the RMSE for the MCMLE to the MPLE for different sample sizes and two different networks, Faux Mesa High and Faux Magnolia High}
\label{rmse}
\end{center}
\end{figure}

\noindent The plots support the fact that larger networks require a larger sample size of simulated networks in order for the MCMLE to outperform the MPLE. While the fairly small Faux Mesa High network only requires a sample size of about 50 networks the larger Faux Magnolia High network already requires a sample size of at least 1500 networks for the MCMLE to surpass the MPLE. These results propose especially for very large networks (e.g. social media data) the question, of how large the sample size has to be set in order to justify the approximately exact, but computationally expensive MCMLE method. The identification of the required sample size is not only becoming more difficult as the size of the network increases, but an increase of the sample size will also extends the mere computation time.

\section*{Bootstrapped MPLE}
As discussed in the previous section, the MPLE converges to the MLE as the size of the network increases. Moreover, the MPLE is able to outperform the MCMLE if the sample size is not chosen large enough. The main reason why the MCMLE is still widely preferred is that in contrast to the MPLE, it does not underestimate the variance of its estimates \cite{vanDuijnetal2009}. By the definition of the ERGM it is obvious that this model is an exponential family distribution where $\theta$ is the natural parameter and $\Gamma (Y)$ is the sufficient statistic. This allows us to work in the framework of exponential family distributions and as a consequence, leads to the conclusion that the sampling distribution of the MLE is multivariate normal with mean vector equal to the MLEs and a covariance matrix equal to the inverse of the negative Hessian matrix $[-H]^{-1}$ of the likelihood function at the MLE. The problem with the MPLE is that calculating $[-H]^{-1}$ by the pseudolikelihood function will underestimate the variance of the MPLE \cite{vanDuijnetal2009}, resulting in an underestimate of the width of the confidence intervals. van Duijn et al. show that constructing 95\% MPLE confidence intervals can result in intervals that only comprise the true value in less than 75\% instead of the demanded 95\%. In this paper, we are going to refer to the MPLE confidence intervals as \textit{logistic regression confidence intervals} simply because the MPLE is calculated using logistic regression methods that also use the inverse of the negative Hessian matrix as an estimate for the covariance matrix. \\[0.3cm]
\begin{figure}[h]
\begin{center}
\includegraphics[scale=.5]{Coverage_Mesa_Magnolia_height05}
\caption{The Coverage Probability results of the Faux Mesa High network (left) and of the Faux Magnolia High network (right) for bootstrapped MPLE, MCMLE and logistic regression }
\label{coverage}
\end{center}
\end{figure}

\begin{figure}[H]
\begin{center}
\includegraphics[scale=.40]{bias_plots_05}
\caption{The boxplots visualize the bias $(\hat{\theta}-\theta)$ over the 500 iterations for the Faux Mesa High network (left) and the Faux Magnolia High network (right).}
\label{bias}
\end{center}
\end{figure}
\noindent Since the MPLE has the advantage of being approximately exact and computationally inexpensive, but has the disadvantage of underestimating corresponding confidence intervals, we are going to apply a technique, referred to as \textit{bootstrapped MPLE}, that was first introduced by Desmarais and Cranmer \cite{Desmarais.2012} and provides a consistent estimate of MPLE confidence intervals. Desmarais and Cranmer argue that this bootstrapped MPLE is a multivariate \textit{M}-estimator \cite{Huber1981}, a special class of robust estimators, meaning that this algorithm consistently estimates the confidence intervals of the MPLE.

\noindent We verify the consistency of the bootstrapped MPLE by conducting a simulation study on the same two networks with the same paramtrization as in the previous chapter: The Faux Mesa High friendship network and the Faux Magnolia High friendship network. \\[0.3cm]
For the simulation study, we determine the MPLE for the model and treat these estimates as the networks' 'true' parameter values. We then use these parameter values to simulate a sample of $1000$ networks from the distribution of $Y$. For each of the $1000$ networks, we calculate 95\% confidence intervals based on the MCMLE and the logistic regression and examine whether the 'true' parameter values lie in these intervals. In addition, we determine the bootstrapped MPLE confidence intervals by sampling $500$ networks for each of the originally sampled $1000$ networks, by using the respective MPLE as parameter values. For every newly sampled network, we again determine the MPLE and then take the $2.5$th and $97.5$th percentile of the $500$ MPLE estimates to obtain a 95\% bootstrap confidence intervals. Similar as for the MCMLE and the logistic regression, we verify whether the 'true' parameter value can be found in the bootstrapped confidence interval.

\noindent Figure \ref{coverage} visualizes the coverage percentages for each of the three methods for both networks. The dashed line is set at $0.95$ and represents the optimal value. It is in evidence that the bootstrapped MPLE performed equally well as the MCMLE, achieving results that obtain the true parameter values in approximately 95\% of the cases. Additionally, a difference in the results between the smaller Faux Mesa High network and larger Faux Magnolia High network is not identifiable. Similar to the results of van Duijn et al. \cite{vanDuijnetal2009} our results for the logistic regression differ distinctively from the anticipated 95\%, visualizing that the MPLE underestimates the variance of its estimates.
Figure \ref{bias} illustrates the bias between the 'true' network coefficients $\theta$ and the MPLE estimates. \textcolor{blue}{(=> not degenerate?, maybe provide boxplots for scnetwork)}
\noindent This simulation study shows that bootstrapped MPLE is able to overcome the main disadvantage of the MPLE by retaining the validity of confidence intervals. Furthermore, this method also adopts the advantage of the MPLE of being consistent. However, the main advantage of the bootstrapped MPLE is the rapidity of the estimation that provides a computational benefit, especially if the size of the examined network is extremely large.

@@

In order to demonstrate the computational advantage of the bootstrapped MPLE to the commonly used MCMLE we apply both approaches to the data on Cosponsorships in the U.S. Senate and U.S. House of Representatives for the 93rd to 108th Congresses developed by Fowler (2006) \cite{Fowler2006a} \cite{Fowler2006b}. The cosponsorship network consists of 2635 nodes and contains information of over 280,000 pieces of legislation proposed in the U.S. House and Senate for the time period 1973 - 2004. In this network bills are tied together based on the similarity of the sets of legislators who cosponsor them. The endogenous statistics we include into this model are the number of edges and the alternating k-star statistic as it was introduced by Snijders et al. \cite{SnijdersTomA.B..2006} and modified by Hunter and Handcock (2006) \cite{Hunter.2006}. The alternating k-star statistic adds one network statistic to the model equal to a weighted alternating sequence of k-star statistics with weight parameter $\lambda$ and is a way to include a networks entire degree distribution as a network statistic. In this model we fix the weight paramter $\lambda=0.4975$. 

@@

Beside these two nodal attributes, we furthermore include the difference between the years of publication as an edge attribute. An edge attribute can be written as a matrix $E$ of the same dimensions as the observed adjacency matrix $G$. In this case entry $e_{ij}$ is filled with the absolute difference between the year of case $i$ and case $j$. Then, the corresponding statistic can be calculated the same way as nodal covariates. Since we do expect a non-linear effect for this edge attribute, we include this variable as a second-degree polynomial by creating a matrix $E^2$, where the matrix gets squared component wise.\\[0.3cm]
We estimate the coefficient of this matrix using both techniques, the MCMLE and the bootstrapped MPLE. The MCMLE requires a sample size of at least $1000$ networks to converge. The bootstrapped MPLE was estimated by using 500 simulated networks. As we described in the chapter {\it Estimation}, only one edge at a time is changed when simulating networks. For better comparison, we chose the same Burn-In ($300.000$ MH-steps) and the same number of iterations ($30.000$ MH-steps) for sampling networks. The results can be found in table \ref{sc_results}.\\[0.3cm]

@@

%\hspace*{-1.5cm}

@@

Edges                                      & -5.657                           & 0.028                             & -4.568                       & 0.038                                    & -5.160 & -4.530  \\ \hline
Spnsor Party                          & 0.592                            & 0.012                             &   1.482                               & 0.009                                   & 1.413 & 1.890  \\ \hline
Alternating k-star                              & 0.253                            & 0.012                             &     0.208                             & 0.015                                   & 0.182 & 0.504  \\ \hline

@@

\label{sc_results}
%\hspace*{-1cm}

@@

\noindent The bootstrapped MPLE is not only simple and fast, it is also embarrassingly parallel, meaning that there is no problem separating the computation into several tasks. In other words, by using multiple cores, the computing time for estimating bootstrapped MPLE confidence intervals can be reduced substantially. Figure \ref{comptime} illustrates the relative computing time of the bootstrapped MPLE using 500 simulated networks and the MCMLE for the three networks Faux Mesa High (205 nodes), Faux Magnolia High (1461 nodes) and Supreme Court Citation (9223 nodes) for an increasing number of computing cores. For the small network we simulate $2000$ networks using a MCMC interval of $2000$ steps, for the medium network we simulate $8000$ networks using a MCMC interval of $5000$ steps and for the large network we simulate 1 million networks using $10,000$ MCMC steps in order to approximate the likelihood appropriately. The chosen sample sizes and MCMC steps are necessary to guarantee a good model fit. We define the simulation time of the bootstrapped MPLE as a function of the number of available computing cores x:

@@

Figure \ref{comptime} demonstrates that the two high school friendship networks only require three cores for the bootstrapped MPLE to outperform the MCMLE. Even more interesting is that the even larger Supreme Court Citation network only requires two cores for the bootstrapped MPLE to outperform the MCMLE.\\[0.3cm]

>>>>

Even though the MCMLE is in general favored over the MPLE there are also cases where the MPLE comes in handy. Foremost, MPLE is quick and simple, since estimation can be done by basic logistic regression and does not exert elaborate MCMC methods. As mentioned in the previous chapter the MPLE approaches the MLE as the size of the networks increase and as a consequence, is a consistent estimator (see Lindsay \cite{Lindsay1988}, Strauss and Ikeda \cite{StraussIkeda1990}, Hyvarinen \cite{Hyvarinen2006}, Desmarais and Cranmer \cite{Desmarais.2012}). This implies that for an increasing number of nodes the MPLE converges in probability to the MLE, meaning that for large enough networks the MPLE becomes an option to computationally intensive MCMC methods.\\

@@

The way we use the term \textit{consistency} in this paper is different and aligns with the way consistency is defined by Lindsay \cite{Lindsay1988}, i.e. instead of considering sub-networks that converge to the full size network, we argue that both, the MLE as well as the MPLE, approach the true coefficient values as the size of networks generally increases. \\%This means that the MPLE of a network on $1000$ nodes is closer to the true coefficient as the MPLE of a network on $100$ nodes. \\
Another major advantage of the MCMLE method is, as shown by van Duijn et al. \cite{vanDuijnetal2009}, that the MCMLE is substantially more efficient than the MPLE, meaning that the variance of the MCMLE is in general lower than the variance of the MPLE. Nevertheless, as shown by Desmarais and Cranmer \cite{Desmarais.2012} the MPLE outperforms the MCMLE if the number of simulated networks used to approximate the likelihood is not chosen large enough. It is even more remarkable that the number of simulated networks needed for the MCMLE, in order to surpass the MPLE increases as the size of the network increases. This means that for very large networks it becomes difficult to determine the number of simulated networks required in order for the MCMLE to outperform the MPLE. In other words, the larger the network of interest is, the larger one has to choose the number of simulated networks in order to justify the time intensive MCMLE approach.\\[0.3cm]
\noindent To demonstrate this disadvantage of the MCMLE we conduct a simulation study using Goodreau's Faux Mesa High School data, which represents a simulation of an in-school friendship network among 203 students as well as the Faux Magnolia High School data, representing an in-school friendship network among 1451 students. The data for both networks originates from Resnick et al. \cite{Resnicketal1997}.\\[0.3cm]
For both networks we first calculate the MCMLE and treat the estimated coefficients as the network's true values $\theta$. Then, we take the same parametrization, using the number of edges, the nodal attribute for gender and the geometrically weighted edgewise shared partners (gwesp) distribution (see Hunter \cite{Hunter.2006}) were we fix the decay parameter $\lambda$ at $0.25$.\\
We simulate $m=500$ new networks using the 'true' coefficients and estimate the MPLE as well as the MCMLE of these simulated networks. For every single simulated network the MCMLE calculation is being repeated several times for $25$ to $10.000$ simulated networks used in the likelihood approximation.

@@

$$RMSE = \sqrt{\sum_{i=1}^{m}(\theta - \hat{\theta}_i)^2}$$

@@

\begin{figure}[h]
\begin{center}
\includegraphics[scale=.5]{RMSE_10k}
\caption{The log of the ratio of the RMSE for the MCMLE to the MPLE for different sample sizes and two different networks, Faux Mesa High and Faux Magnolia High}
\label{rmse}
\end{center}
\end{figure}

\noindent The plots support the fact that larger networks require a larger sample size of simulated networks for the MCMLE to outperform the MPLE. While the fairly small Faux Mesa High network only requires a sample size of about $50-100$ networks the larger Faux Magnolia High network already requires a sample size of at least 1500 networks for the MCMLE to surpass the MPLE. These results propose especially for very large networks (e.g. social media data) the question, of how large the sample size has to be set in order to justify the approximately exact, but computationally expensive MCMLE method. The identification of the required sample size is not only becoming more difficult as the size of the network increases, but an increase of the sample size will also extend the mere computation time.

\section*{Bootstrapped MPLE}
As discussed in the previous section, the MPLE converges to the MLE as the size of the network increases. Moreover, the MPLE is able to outperform the MCMLE if the sample size is not chosen large enough. The main reason why the MCMLE is still widely preferred is that in contrast to the MPLE, it does not underestimate the standard deviation (van Duijn et al. \cite{vanDuijnetal2009}). By the definition of the ERGM it is obvious that this model is an exponential family distribution where $\theta$ is the natural parameter and $\Gamma (Y)$ is the sufficient statistic. For exponential family distributions it is known that the sampling distribution of the MLE is multivariate normal with mean vector equal to the MLE and a covariance matrix equal to the inverse of the negative Hessian matrix $[-H]^{-1}$ of the likelihood function at the MLE. The problem with the MPLE is that calculating $[-H]^{-1}$ by the pseudolikelihood function will underestimate the variance of the MPLE \cite{vanDuijnetal2009}, resulting in an underestimate of the width of the confidence intervals. van Duijn et al. show that constructing 95\% MPLE confidence intervals can result in intervals that comprise the true value in less than 75\% instead of the demanded 95\%. In this paper, we are going to refer to the MPLE confidence intervals as \textit{logistic regression confidence intervals} simply because the MPLE is calculated using logistic regression methods that also use the inverse of the negative Hessian matrix as an estimate for the covariance matrix. \\[0.3cm]
\noindent Since the MPLE has the advantage of being approximately exact and computationally inexpensive, but has the disadvantage of underestimating corresponding confidence intervals, we apply a technique referred to as \textit{bootstrapped MPLE}, that was first introduced by Desmarais and Cranmer \cite{Desmarais.2012} and provides a consistent estimate of MPLE confidence intervals. Desmarais and Cranmer argue that this bootstrapped MPLE is a multivariate \textit{M}-estimator (see Huber \cite{Huber1981}), a special class of robust estimators, meaning that this algorithm consistently estimates the confidence intervals of the MPLE.

\noindent We verify the consistency of the bootstrapped MPLE by conducting a simulation study on the same two networks with the same parametrization as in the previous chapter: The Faux Mesa High friendship network and the Faux Magnolia High friendship network. \\[0.3cm]
For the simulation study, we determine the MPLE for the model and treat these estimates as the networks' 'true' parameter values. We then use these parameter values to simulate a sample of $1000$ networks from the distribution of $Y$. For each of the $1000$ networks, we calculate 95\% confidence intervals based on the MCMLE and the logistic regression and examine whether the 'true' parameter values lie in these intervals. In addition, we determine the bootstrapped MPLE confidence intervals by sampling $500$ networks for each of the originally sampled $1000$ networks, by using the respective MPLE as parameter values. For every newly sampled network, we again determine the MPLE and then take the $2.5$th and $97.5$th percentile of the $500$ MPLE estimates to obtain a 95\% bootstrap confidence intervals. Similar as for the MCMLE and the logistic regression, we verify whether the 'true' parameter value can be found in the bootstrapped confidence interval.

\begin{figure}[]
\begin{center}
\includegraphics[scale=.5]{Coverage_Mesa_Magnolia_height05}
\caption{The Coverage Probability results of the Faux Mesa High network (left) and of the Faux Magnolia High network (right) for bootstrapped MPLE, MCMLE and logistic regression }
\label{coverage}
\end{center}
\end{figure}

\begin{figure}[]
\begin{center}
\includegraphics[scale=.40]{bias_plots_05}
\caption{The boxplots visualize the bias $(\hat{\theta}-\theta)$ over the 500 iterations for the Faux Mesa High network (left) and the Faux Magnolia High network (right).}
\label{bias}
\end{center}
\end{figure}

\noindent Figure \ref{coverage} visualizes the coverage percentages for each of the three methods for both networks. The dashed line is set at $0.95$ and represents the optimal value. It is evident that the bootstrapped MPLE performed equally well as the MCMLE, achieving results that obtain the true parameter values in approximately 95\% of the cases. Additionally, a difference in the results between the smaller Faux Mesa High network and larger Faux Magnolia High network is not identifiable. Similar to the results of van Duijn et al. \cite{vanDuijnetal2009} our results for the logistic regression differ distinctively from the anticipated 95\%, visualizing that the MPLE underestimates the variance of its estimates.
Figure \ref{bias} illustrates the bias between the 'true' network coefficients $\theta$ and the MPLE estimates. We can abstract from this figure that the median MPLE estimates concur with the networks 'true' coefficients.\\
\noindent This simulation study shows that bootstrapped MPLE is able to overcome the main disadvantage of the MPLE by retaining the validity of confidence intervals. Furthermore, this method also adopts the advantage of the MPLE of being consistent. However, the main advantage of the bootstrapped MPLE is the rapidity of the estimation that provides a computational benefit, especially if the size of the examined network is extremely large.

@@

To demonstrate the computational advantage of the bootstrapped MPLE to the commonly used MCMLE we apply both approaches to the data on Cosponsorships in the U.S. Senate and U.S. House of Representatives for the 93rd to 108th Congresses developed by Fowler (2006) \cite{Fowler2006a} \cite{Fowler2006b}. The cosponsorship network consists of 2635 nodes and contains information of over 280,000 pieces of legislation proposed in the U.S. House and Senate for the time period 1973 - 2004. In this network bills are tied together based on the similarity of the sets of legislators who cosponsor them. The endogenous statistics we include into this model are the number of edges and the alternating k-star statistic as it was introduced by Snijders et al. \cite{SnijdersTomA.B..2006} and modified by Hunter and Handcock (2006) \cite{Hunter.2006}. The alternating k-star statistic adds one network statistic to the model equal to a weighted alternating sequence of k-star statistics with weight parameter $\lambda$ and is a way to include a networks entire degree distribution as a network statistic. In this model we fix the weight paramter $\lambda=0.4975$. 

@@

%Beside these two nodal attributes, we furthermore include the difference between the years of publication as an edge attribute. An edge attribute can be written as a matrix $E$ of the same dimensions as the observed adjacency matrix $G$. In this case entry $e_{ij}$ is filled with the absolute difference between the year of case $i$ and case $j$. Then, the corresponding statistic can be calculated the same way as nodal covariates. Since we do expect a non-linear effect for this edge attribute, we include this variable as a second-degree polynomial by creating a matrix $E^2$, where the matrix gets squared component wise.\\[0.3cm]
We estimate the coefficient of this matrix using both techniques, the MCMLE and the bootstrapped MPLE. The MCMLE requires a sample size of at least $1000$ networks to converge. The bootstrapped MPLE was estimated by using 500 simulated networks. As we described in the chapter {\it Estimation}, only one edge at a time is changed when simulating networks. For better comparison, we chose the same Burn-In ($300.000$ MH-steps) and the same number of iterations ($30.000$ MH-steps) for sampling networks. The results can be found in table \ref{co_results}.\\
It is interesting to see that the standard error calculated by the logistic regression approach is constantly smaller than the standard error of the MCMLE, which leads to inaccurate confidence intervals as shown in figure \ref{coverage}. Furthermore, we can observe that the coefficients for the number of edges and the nodal attribute are very similar, but the estimates for the alternating k-star statistic do differ \textcolor{blue}{reason?}. The bootstrapped MPLE bounds, however, include each estimate for both methods.  \\[0.3cm]

@@

\hspace*{-1.5cm}

@@

Edges                                      & -5.884 & 0.065                             & -5.869                       & 0.015                                    & -6.007 & -5.751  \\ \hline
Sponsor Party                          & 1.440                            & 0.015                             &   1.440                               & 0.015                                   & 1.411 & 1.467  \\ \hline
Alternating k-star                              & 0.125                            & 0.064                             &     0.108                             & 0.006                                   & -0.011 & 0.2379  \\ \hline

@@

\label{co_results}
\hspace*{-1cm}

@@

\noindent The bootstrapped MPLE is not only simple and fast, it is also embarrassingly parallel, meaning that there is no problem separating the computation into several tasks. In other words, by using multiple cores, the computing time for estimating bootstrapped MPLE confidence intervals can be reduced substantially. Figure \ref{comptime} illustrates the relative computing time of the bootstrapped MPLE using 500 simulated networks and the MCMLE for the three networks Faux Mesa High (205 nodes), Faux Magnolia High (1461 nodes) and Cosponsorship (2635 nodes) for an increasing number of computing cores. For the small network we simulate $2000$ networks using a MCMC interval of $2000$ steps, for the medium network we simulate $8000$ networks using a MCMC interval of $5000$ steps and for the large network we simulate $10000$ networks using $30,000$ MCMC steps in order to approximate the likelihood appropriately. The chosen sample sizes and MCMC steps are necessary to guarantee a good model fit. We define the simulation time of the bootstrapped MPLE as a function of the number of available computing cores x:

@@

Figure \ref{comptime} demonstrates that all three networks only require two cores for the bootstrapped MPLE to outperform the computing time of the MCMLE and that the computing time can further be reduced if more computing cores are available. This figure also depicts that larger network in general require a longer computation time and will benefit more if the bootstrapped MPLE is used.\\ 
Of course the actual computing time for a network always depends on the statistics that are included in the network, but in general larger networks require longer computation times, since a larger MCMC sample size is required and more MC steps are necessary to simulate a new network that does not overly depend on the previous sample. This makes the bootstrapped MPLE a beneficial alternative especially for very large networks. \\[0.3cm]
